---
title: "Exc6 Model selection"
author: "Henrik Sausen"
date: "2023-05-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Problem 2 

```{r}
library(ISLR)
library(GGally)
library(ggplot2)

ggpairs(subset(Credit, select = -c(ID,Gender, Student, Married, Ethnicity))) + theme_minimal()

```


#Problem 3


Best subset selection based on Cp, BIC and R^2 adjusted.
```{r}
library(leaps)

credit.data <- subset(Credit, select = -ID)

#using 75% as training and 25% as test 
set.seed(1)
train_index <- sample(1:nrow(credit.data), nrow(credit.data)*0.75)
credit.data.training <- credit.data[train_index,]
credit.data.test <- credit.data[-train_index,]

best.sub.select <- regsubsets(Balance ~., data = credit.data.training) #all predictors except ID

r2adj <- summary(best.sub.select)$adjr2
cp <- summary(best.sub.select)$cp
bic <- summary(best.sub.select)$bic
rss <- summary(best.sub.select)$rss
```


Plots
```{r}
par(mfrow=c(2,2))

plot(rss, type = 'l', xlab = "number of variables")
rss.min <- which.min(rss)
points(rss.min, rss[rss.min], col = 'red')

plot(cp, type = 'l', xlab = "number of variables")
cp.min <- which.min(cp)
points(cp.min, cp[cp.min], col = 'red')

plot(r2adj, type = 'l', xlab = "number of variables")
r2adj.max <- which.max(r2adj)
points(r2adj.max, r2adj[r2adj.max], col = 'red')

plot(bic, type = 'l', xlab = "number of variables")
bic.min <- which.min(bic)
points(bic.min, bic[bic.min], col = 'red')
```




```{r}
predict.regsubsets=function(object,newdata,id,...){
  form=as.formula(object$call[[2]])
  mat=model.matrix(form,newdata)
  coefi=coef(object,id=id)
  xvars=names(coefi)
  mat[,xvars]%*%coefi
}
```



Best subset selection using 10-fold CV.
```{r}
num.predictors <- 10 #number of predictors we have in model to start with. 

k <- 10
n.training <- nrow(credit.data.training) 


set.seed(1)
folds <- sample(rep(1:k, length = n.training), replace = T) #list of length n, with numbers between 1 and k = 10. We assign one of each observation to one of the folds 1..10
cv.errors <- matrix(NA, k, num.predictors) #each row indicates a fold, and column inidicates a predictor 

for (j in 1:k) {
  best.fit <- regsubsets(Balance ~., data = credit.data.training[folds != j, ], nvmax = num.predictors) #all rows (observations) that do not equal j (that is not in the j'th fold)
  for (i in 1:num.predictors) {
    pred <- predict(best.fit, credit.data.training[folds == j, ], id = i) #here we make predictions for the observations in the current fold j, id-> each          subset size i 
    cv.errors[j,i] <- mean((credit.data.training$Balance[folds == j] - pred)^2) #calculate the mean of the error of the predicted model 
  }
}

#we want the smallest average error over all the folds for each predictor. ¨

mse.cv <- apply(cv.errors, 2, mean) #2 indicates that we calculate mean by column, 1 means by row
plot(mse.cv, pch = num.predictors, type = 'b', xlab = 'number of predictors')
```

```{r}
optimal_num_pred <- which.min(rmse.cv)

reg.best <- regsubsets(Balance ~., data = credit.data.training)
coef(reg.best, 4) #the predictors below are the model with 4 predictors that is the best, chosen by best subset selection. 
```


```{r}
model <- lm(Balance ~ Income + Limit + Cards + Student, data = credit.data.training)
model.predict <- predict(model, newdata = credit.data.test)
summary(model)
```

Now we want to calculate the error of our prediction with the best model - that we found above. We calculate the test Mean Square Error below: 

```{r}
MSE = mean((credit.data$Balance[-train_index]-model.predict)^2)
MSE
```

#Problem 4

Now we want to select the best model using FW, BW and Hybrid Stepwise Selection. 
```{r}
#forward 
regfit.fw <- regsubsets(Balance ~.,data = credit.data.training, method = "forward")

bic_fw <- summary(regfit.fw)$bic
cp_fw <- summary(regfit.fw)$cp
radj_fw <- summary(regfit.fw)$adjr2
rss_fw <- summary(regfit.fw)$rss

par(mfrow = c(2,2))

plot(rss_fw, type = 'l', xlab = "number of variables")
rss_fw.min <- which.min(rss_fw)
points(rss_fw.min, rss_fw[rss_fw.min], col = 'red')

plot(cp_fw, type = 'l', xlab = "number of variables")
cp_fw.min <- which.min(cp_fw)
points(cp_fw.min, cp_fw[cp_fw.min], col = 'red')

plot(radj_fw, type = 'l', xlab = "number of variables")
radj_fw.max <- which.max(radj_fw)
points(radj_fw.max, radj_fw[radj_fw.max], col = 'red')

plot(bic_fw, type = 'l', xlab = "number of variables")
bic_fw.min <- which.min(bic_fw)
points(bic_fw.min, bic_fw[bic_fw.min], col = 'red')
```


#Problem 5 Ridge Regression

```{r}
library(glmnet)

#create model matrix 
X <- model.matrix(Balance ~., data = credit.data)[,-1]
Y <- credit.data$Balance

set.seed(1)

Ridge.model <- cv.glmnet(X[train_index,],Y[train_index], alpha = 0) #uses cv to find the optimal lambda
coef(Ridge.model)
```

```{r}
lm.model <- lm(Balance ~., data = credit.data[train_index,])
summary(lm.model)

lm.predict <- predict(lm.model, newdata = credit.data.test)

MSE_lm <- mean((credit.data.test$Balance - lm.predict)^2)

MSE_lm
```
Comparing these two, we see that the estimated parameters are smaller for the Ridge regression. 

Plots:
```{r}

set.seed(1)
best.lambda <- Ridge.model$lambda.min

ridge.prediction <- predict(Ridge.model, s = best.lambda, newx = X[-train_index,])

MSE_best_lambd <- mean((Y[-train_index] - ridge.prediction)^2)

MSE_best_lambd
```

Still less mean square error than for the model selection done above. But Ridge is poweful, when we have Least Square estimates with high variance. Then Ridge will trade a large decrease in variance for a small increase in bias. This happens because an increased value of lambda will reduce the complexity of the model, meaning that the coefficient estimates decrease in value. Since test-MSE is closely related to both variance and bias, the minimal test-MSE of the ridge regression will give a good trade of variance and bias in the model. However, we see that the MSE is more for the Ridge regression than for the normal linear model in this case.

```{r}
plot(Ridge.model)
```

#Problem 5 Lasso 

```{r}
set.seed(1) #such that we get the same lambda using cross validation
lasso.model <- cv.glmnet(X[train_index, ], Y[train_index], alpha = 1)

best_lambda_lass <- lasso.model$lambda.min

lass_predict <- predict(lasso.model, s = best_lambda_lass, newx = X[-train_index,])

MSE_lass <- mean((credit.data.test$Balance - lass_predict)^2)
MSE_lass
```
```{r}
coef(lasso.model)
```
We see that the Lasso sets some parameters to 0. This happens because of the l1 penalisation. 


```{r}
plot(lasso.model)
```
#Problem 7 Principal components
How many PC's should we use for this dataset? We use cross-validation in order to decide. 

```{r}
library(pls)
set.seed(1)
pcr.model <- pcr(Balance ~., data = credit.data.training, scale = T, validation = 'CV')
validationplot(pcr.model, val.type = "MSEP")
```
The plot shows the root-mean square error. 
```{r}
summary(pcr.model)
```

We see that the amount of extra variance explained after 8 components is quite small. Hence, 8 PCs would suffice. 

#Problem 8 PCR 
```{r}
pcr.predict <- predict(pcr.model, newdata = credit.data.test, ncomp = 10)

pcr_MSE <- mean((credit.data.test$Balance - pcr.predict)^2)

pcr_MSE
```

Using 10 PCs we get the MSE above in the test set. 
